---
title: "Analysis"
output: pdf_document
---

```{r}
# Model Data

library(DBI)
library(RMySQL)
library(dplyr)
library(tidyverse)

con <- dbConnect(
  RMySQL::MySQL(),
  dbname = "qbs181",
  host = "localhost",
  user = "michael",
  password = "431679"
)

model_data <- dbGetQuery(con, "
  SELECT
    id,
    neighbourhood_group,
    cancellation_policy,
    room_type,
    construction_year,
    price AS price_numeric,
    LOG(price) AS log_price,
    minimum_nights,
    number_of_reviews,
    availability_365,
    service_fee,
    reviews_per_month,
    GEOID_x,
    median_income,
    population,
    median_rent
  FROM final_data
  WHERE price > 0
    AND median_income IS NOT NULL
    AND room_type IS NOT NULL
    AND neighbourhood_group IS NOT NULL;
")

nrow(model_data)
glimpse(model_data)

write_csv(model_data, "data/model_data.csv")
```

```{r}
# Custom Function 1: Run model diagnostics (VIF, residual plots)
run_model_diagnostics <- function(model) {
  print("--- VIF (Multicollinearity) ---")
  if (length(coef(model)) > 2) { # VIF requires at least two predictor variables
    print(car::vif(model))
  } else {
    print("The model has too few variables to calculate VIF.")
  }
  
  print("--- Residual Plots ---")
  par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid
  plot(model)           # Plot residual diagnostics
  par(mfrow = c(1, 1))  # Reset plotting layout
}
```

```{r}
# 1. Multiple Linear Regression (MLR)

mlr_model <- lm(
  log_price ~ room_type + neighbourhood_group + minimum_nights +
              number_of_reviews + availability_365 + 
              median_income + median_rent + cancellation_policy,
  data = model_data
)

# Display the model summary (coefficients, R-squared, p-values, etc.)
summary(mlr_model)

# 2. Run the custom diagnostic function
run_model_diagnostics(mlr_model)

```

Model Results and Diagnostics

The multiple linear regression model was fitted to explain variation in the log-transformed Airbnb price (log_price) using room type, neighborhood group, stay characteristics, and neighborhood-level socioeconomic indicators.
The model’s explanatory power is extremely low (R² = 0.0003), suggesting that only a negligible fraction of price variability is captured by the selected predictors. Among all variables, only median rent is statistically significant (p ≈ 0.037), showing a weak negative association with price, while median income, minimum nights, and other factors are not significant.

Variance Inflation Factors (VIFs) reveal moderate multicollinearity between median_income (VIF ≈ 8.1) and median_rent (VIF ≈ 7.8), indicating that these two variables capture overlapping regional effects and one should be removed in future models.

Residual diagnostics show that linearity is approximately satisfied, but the residuals deviate from normality (S-shaped Q-Q plot) and exhibit mild heteroskedasticity (Scale-Location plot). A few high-leverage observations were detected but are not severe.
Overall, the model performs poorly in predictive terms, implying that key structural or locational factors are missing. Future work should include more detailed spatial and property-level variables to improve explanatory power and model fit.

```{r}
nyc_acs_wide <- nyc_acs_data_transformed %>%
  select(GEOID = GEOID.x, NAME = NAME.x, variable, estimate, geometry) %>%
  pivot_wider(
    names_from = variable,
    values_from = estimate,
    values_fn = mean
  )

plot_income_map <- function(nyc_acs_wide) {
  ggplot(nyc_acs_wide) +
    geom_sf(aes(fill = median_income), color = NA) +
    scale_fill_viridis_c(
      option = "plasma",
      labels = scales::dollar,
      na.value = "grey90"
    ) +
    labs(
      title = "Median Household Income by ZIP Code (NYC, 2020)",
      fill = "Median Income ($)"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "right"
    )
}

plot_income_map(nyc_acs_wide)
```

```{r}
model_data <- model_data %>%
  left_join(
    df_cleaned_first %>% select(id, lat, long),
    by = "id"
  )

plot_airbnb_prices <- function(model_data) {
  ggplot(data = model_data %>% filter(price_numeric < 1000)) +
    geom_point(
      aes(x = long, y = lat, color = price_numeric),
      size = 0.7,
      alpha = 0.6
    ) +
    scale_color_viridis_c(
      option = "inferno",
      trans = "log",
      name = "Price (log scale)",
      labels = scales::dollar
    ) +
    coord_equal() +
    labs(
      title = "NYC Airbnb Price Distribution (< $1000 Listings)",
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "right"
    )
}

plot_airbnb_prices(model_data)
```

(a) Median Household Income by ZIP Code (NYC, 2020)

The first choropleth map visualizes the spatial distribution of median household income across ZIP Code Tabulation Areas (ZCTAs) in New York City.
High-income neighborhoods are primarily concentrated in central and lower Manhattan, as well as parts of northwest Brooklyn.
By contrast, the Bronx and southern Queens exhibit substantially lower income levels.
This socioeconomic gradient provides essential context for understanding local disparities in housing affordability and Airbnb market segmentation.

(b) Airbnb Price Distribution (< $1000 Listings)

The second map plots individual Airbnb listings under $1000 per night, colored by price on a logarithmic scale.
Listings are densely clustered in Manhattan and northern Brooklyn, particularly in touristic and central business districts.
The highest prices (bright yellow points) are found in Midtown and Downtown Manhattan, reflecting premium demand and location-based price effects.

(c) Combined Interpretation

Together, these maps highlight a spatial alignment between economic prosperity and Airbnb pricing.
Higher-income neighborhoods also tend to host higher-priced listings, suggesting that income and geographic desirability jointly influence market outcomes.