---
title: "Custom Functions Library"
author: "QBS181 Data Wrangling Project"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Custom Functions Library

This Rmd file contains all custom functions used in the project.

## 1.Data Cleaning Functions

### 1.1 Cleaning airbnb main functions

```{r clean-airbnb-main}
#' Clean raw Airbnb data
#' 
#' @description Performs comprehensive cleaning on Airbnb data, including standardizing column names, type conversion, missing value handling, etc.
#' @param data Raw Airbnb dataframe
#' @param file_path Optional, CSV file path (if reading from file)
#' @return Cleaned dataframe
#' @examples
#' df_cleaned <- clean_airbnb_data(raw_data)
clean_airbnb_data <- function(data = NULL, file_path = NULL) {
  # Load necessary libraries
  library(tidyverse)
  library(janitor)
  library(lubridate)
  
  # If file path is provided, read from file
  if (!is.null(file_path)) {
    data <- read_csv(file_path)
  }
  
  # Standardize column names
  data <- data %>% clean_names()
  
  # Type conversion
  data <- data %>%
    mutate(
      price = as.numeric(price),
      service_fee = as.numeric(service_fee)
    )
  
  # Handle missing values
  data <- handle_missing_values(data)
  
  # Clean date column
  data <- clean_date_column(data, date_col = "last_review")
  
  # Clean minimum_nights column
  data <- clean_minimum_nights(data)
  
  return(data)
}
```

### 1.2 Handle Missing Values

```{r handle-missing-values}
#' Fill missing values with median
#' 
#' @description Fills missing values in specified numeric columns using the median
#' @param data Dataframe
#' @param cols Vector of column names to process, defaults to reviews_per_month and availability_365
#' @return Processed dataframe
#' @examples
#' df <- handle_missing_values(df, cols = c("reviews_per_month", "availability_365"))
handle_missing_values <- function(data, cols = c("reviews_per_month", "availability_365")) {
  for (col in cols) {
    if (col %in% names(data)) {
      median_val <- median(data[[col]], na.rm = TRUE)
      data[[col]] <- replace_na(data[[col]], median_val)
    }
  }
  return(data)
}
```

### 1.3 Clean Date Column

```{r clean-date-column}
#' Clean and standardize date column
#' 
#' @description Converts date column to standard format and handles abnormal date values
#' @param data Dataframe
#' @param date_col Date column name
#' @param date_limit Date upper limit, default "2021-01-01"
#' @return Processed dataframe
#' @examples
#' df <- clean_date_column(df, date_col = "last_review")
clean_date_column <- function(data, date_col = "last_review", date_limit = "2021-01-01") {
  library(lubridate)
  
  if (!date_col %in% names(data)) {
    warning(paste("Column", date_col, "not found in data"))
    return(data)
  }
  
  # Convert to date format
  data[[date_col]] <- mdy(data[[date_col]])
  
  # Calculate median of valid dates
  date_limit_date <- as.Date(date_limit)
  valid_dates <- data[[date_col]][!is.na(data[[date_col]]) & 
                                   data[[date_col]] <= date_limit_date]
  median_date <- median(valid_dates, na.rm = TRUE)
  
  # Replace abnormal dates
  data <- data %>%
    mutate(!!sym(date_col) := if_else(
      is.na(!!sym(date_col)) | !!sym(date_col) > date_limit_date,
      median_date,
      !!sym(date_col)
    ))
  
  return(data)
}
```

### 1.4 Clean minimum_nights Column

```{r clean-minimum-nights}
#' Clean minimum_nights column, handle outliers
#' 
#' @description Restricts minimum_nights to a reasonable range (1-365)
#' @param data Dataframe
#' @param min_val Minimum value, default 1
#' @param max_val Maximum value, default 365
#' @return Processed dataframe
#' @examples
#' df <- clean_minimum_nights(df)
clean_minimum_nights <- function(data, min_val = 1, max_val = 365) {
  if (!"minimum_nights" %in% names(data)) {
    warning("Column 'minimum_nights' not found in data")
    return(data)
  }
  
  data <- data %>%
    mutate(
      minimum_nights = case_when(
        minimum_nights <= 0 ~ min_val,
        minimum_nights > max_val ~ max_val,
        TRUE ~ minimum_nights
      )
    )
  
  return(data)
}
```

## 2. Visualization Functions

### 2.1 Plot Income Map

```{r plot-income-map}
#' Plot map of median income by NYC ZIP code
#' 
#' @description Creates a choropleth map of income distribution using ggplot2 and sf
#' @param nyc_acs_wide sf object containing ACS data and geometry information
#' @param income_col Income column name, default "median_income"
#' @param title Chart title
#' @return ggplot object
#' @examples
#' plot_income_map(nyc_acs_wide)
plot_income_map <- function(nyc_acs_wide, 
                            income_col = "median_income",
                            title = "Median Household Income by ZIP Code (NYC, 2020)") {
  library(ggplot2)
  library(sf)
  library(viridis)
  library(scales)
  
  ggplot(nyc_acs_wide) +
    geom_sf(aes(fill = !!sym(income_col)), color = NA) +
    scale_fill_viridis_c(
      option = "plasma",
      labels = dollar,
      na.value = "grey90"
    ) +
    labs(
      title = title,
      fill = "Median Income ($)"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "right"
    )
}
```

### 2.2 Plot scatter of Airbnb listing prices

```{r plot-airbnb-prices}
#' Plot scatter map of Airbnb listing prices
#' 
#' @description Displays location and price of Airbnb listings on a map
#' @param model_data Dataframe containing listing data, must include lat, long, price_numeric columns
#' @param max_price Maximum display price, default 1000
#' @param title Chart title
#' @return ggplot object
#' @examples
#' plot_airbnb_prices(model_data, max_price = 1000)
plot_airbnb_prices <- function(model_data, 
                               max_price = 1000,
                               title = "NYC Airbnb Price Distribution (< $1000 Listings)") {
  library(ggplot2)
  library(viridis)
  library(scales)
  
  ggplot(data = model_data %>% filter(price_numeric < max_price)) +
    geom_point(
      aes(x = long, y = lat, color = price_numeric),
      size = 0.7,
      alpha = 0.6
    ) +
    scale_color_viridis_c(
      option = "inferno",
      trans = "log",
      name = "Price (log scale)",
      labels = dollar
    ) +
    coord_equal()
```

### 2.3 Plot Borough Distribution

```{r plot-borough-distribution}
#' Plot the distribution of listings by borough
#' 
#' @description Create a bar chart showing the number of listings in each borough
#' @param data Dataframe containing neighbourhood_group column
#' @param title Chart title
#' @return ggplot object
#' @examples
#' plot_borough_distribution(data_cleaned)
plot_borough_distribution <- function(data, 
                                      title = "NYC Airbnb Listing Count by Borough") {
  library(ggplot2)
  library(forcats)
  library(dplyr)
  
  borough_data <- data %>%
    count(neighbourhood_group, name = "n") %>%
    mutate(
      neighbourhood_group = fct_reorder(neighbourhood_group, n, .desc = TRUE)
    )
  
  ggplot(borough_data, aes(x = neighbourhood_group, y = n)) +
    geom_col(fill = "#56B4E9") +
    labs(
      title = title,
      subtitle = "Manhattan and Brooklyn are the dominant boroughs",
      x = "Borough",
      y = "Number of Listings",
      caption = "Source: NYC Airbnb Data (2019)"
    ) +
    theme_minimal(base_family = "Arial") +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

### 2.4 Plot Room Type Distribution

```{r plot-room-type-distribution}
#' Plot room type distribution
#' 
#' @description Create a bar chart or stacked bar chart showing room type distribution
#' @param data Dataframe containing room_type column
#' @param by_borough Whether to group by borough, default FALSE
#' @param title Chart title
#' @return ggplot object
#' @examples
#' plot_room_type_distribution(data_cleaned)
#' plot_room_type_distribution(data_cleaned, by_borough = TRUE)
plot_room_type_distribution <- function(data, 
                                        by_borough = FALSE,
                                        title = "Room Type Distribution") {
  library(ggplot2)
  library(dplyr)
  library(viridis)
  
  if (by_borough) {
    # Stacked bar chart
    p <- ggplot(data, aes(x = neighbourhood_group, fill = room_type)) +
      geom_bar(position = "stack") +
      scale_fill_viridis_d(option = "cividis") +
      labs(
        title = "Room Type Composition by Borough",
        x = "Borough",
        y = "Number of Listings",
        fill = "Room Type"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
      )
  } else {
    # Simple bar chart
    p <- ggplot(data, aes(x = room_type)) +
      geom_bar(fill = "darkorange") +
      labs(
        title = title,
        x = "Room Type",
        y = "Count"
      ) +
      theme_minimal()
  }
  
  return(p)
}
```

## 3. Modeling functions

### 3.1 Model Diagnostics Function

```{r run-model-diagnostics}
#' Run regression model diagnostics
#' 
#' @description Calculates VIF (Variance Inflation Factor) and plots residual diagnostic plots
#' @param model Linear regression model object (lm)
#' @param plot_residuals Whether to plot residuals, default TRUE
#' @return No return value, prints results and charts directly
#' @examples
#' model <- lm(log_price ~ room_type + neighbourhood_group, data = model_data)
#' run_model_diagnostics(model)
run_model_diagnostics <- function(model, plot_residuals = TRUE) {
  library(car)
  
  cat("--- VIF (Multicollinearity) ---\n")
  if (length(coef(model)) > 2) {
    vif_results <- car::vif(model)
    print(vif_results)
  } else {
    cat("The model has too few variables to calculate VIF.\n")
  }
  
  if (plot_residuals) {
    cat("\n--- Residual Plots ---\n")
    par(mfrow = c(2, 2))
    plot(model)
    par(mfrow = c(1, 1))
  }
}
```

### 3.2 Prepare Modeling Data

```{r prepare-model-data}
#' Prepare modeling data from database or CSV file
#' 
#' @description Reads data required for modeling from MySQL database or CSV file
#' @param source Data source type: "database" or "csv"
#' @param db_connection Database connection object (if source = "database")
#' @param csv_path CSV file path (if source = "csv")
#' @param table_name Database table name, default "final_data"
#' @return Modeling dataframe
#' @examples
#' # Read from database
#' con <- dbConnect(RMySQL::MySQL(), dbname = "qbs181", ...)
#' model_data <- prepare_model_data("database", db_connection = con)
#' 
#' # Read from CSV
#' model_data <- prepare_model_data("csv", csv_path = "data/model_data.csv")
prepare_model_data <- function(source = "database",
                               db_connection = NULL,
                               csv_path = "data/model_data.csv",
                               table_name = "final_data") {
  if (source == "database") {
    if (is.null(db_connection)) {
      stop("Database connection is required when source = 'database'")
    }
    
    query <- paste("
      SELECT
        id,
        neighbourhood_group,
        cancellation_policy,
        room_type,
        construction_year,
        price AS price_numeric,
        LOG(price) AS log_price,
        minimum_nights,
        number_of_reviews,
        availability_365,
        service_fee,
        reviews_per_month,
        GEOID_x,
        median_income,
        population,
        median_rent
      FROM", table_name, "
      WHERE price > 0
        AND median_income IS NOT NULL
        AND room_type IS NOT NULL
        AND neighbourhood_group IS NOT NULL;
    ")
    
    model_data <- dbGetQuery(db_connection, query)
    
  } else if (source == "csv") {
    model_data <- read_csv(csv_path, show_col_types = FALSE)
  } else {
    stop("source must be either 'database' or 'csv'")
  }
  
  return(model_data)
}
```

## 4. Spatial Analysis Functions

### 4.1 Spatial Join Airbnb and ACS Data

```{r spatial-join-airbnb-acs}
#' Perform spatial join of Airbnb and ACS data
#' 
#' @description Merges Airbnb listing point data with ACS regional data via spatial join
#' @param airbnb_data Airbnb dataframe, must contain lat and long columns
#' @param acs_data ACS sf object
#' @param crs Coordinate reference system, default 4326 (WGS84)
#' @param drop_geometry Whether to drop geometry info, default TRUE
#' @return Joined dataframe
#' @examples
#' joined_data <- spatial_join_airbnb_acs(airbnb_data, nyc_acs_data)
spatial_join_airbnb_acs <- function(airbnb_data, 
                                    acs_data,
                                    crs = 4326,
                                    drop_geometry = TRUE) {
  library(sf)
  library(dplyr)
  
  # Ensure Airbnb data has coordinates
  airbnb_data <- airbnb_data %>%
    filter(!is.na(long) & !is.na(lat))
  
  # Convert to sf object
  airbnb_sf <- airbnb_data %>%
    st_as_sf(coords = c("long", "lat"), crs = crs)
  
  # Ensure ACS data CRS matches
  if (st_crs(acs_data)$epsg != crs) {
    acs_data <- st_transform(acs_data, crs = crs)
  }
  
  # Perform spatial join
  joined_data <- st_join(airbnb_sf, acs_data, join = st_intersects)
  
  # Drop geometry if requested
  if (drop_geometry) {
    joined_data <- joined_data %>%
      st_drop_geometry() %>%
      as.data.frame()
  }
  
  return(joined_data)
}
```

### 4.2 Create Wide Table Format Data

```{r create-wide-table}
#' Convert long format ACS data to wide format
#' 
#' @description Converts long format data containing variable and estimate columns to wide format
#' @param long_data Long format dataframe, must contain variable and estimate columns
#' @param id_cols Vector of column names to use as ID
#' @param value_fn Aggregation function, default mean
#' @return Wide format dataframe
#' @examples
#' wide_data <- create_wide_table(long_data, id_cols = c("id", "GEOID"))
create_wide_table <- function(long_data, 
                              id_cols = "id",
                              value_fn = mean) {
  library(tidyr)
  library(dplyr)
  
  wide_data <- long_data %>%
    pivot_wider(
      names_from = variable,
      values_from = estimate,
      id_cols = all_of(id_cols),
      values_fn = value_fn
    )
  
  return(wide_data)
}
```

## 5. Utility Functions

### 5.1 Check Data Quality

```{r check-data-quality}
#' Check data quality
#' 
#' @description Checks dataframe for quality issues such as missing values and duplicates
#' @param data Dataframe
#' @param print_summary Whether to print summary, default TRUE
#' @return List containing quality check results
#' @examples
#' quality_report <- check_data_quality(data)
check_data_quality <- function(data, print_summary = TRUE) {
  library(dplyr)
  
  results <- list(
    n_rows = nrow(data),
    n_cols = ncol(data),
    missing_values = colSums(is.na(data)),
    duplicate_rows = sum(duplicated(data)),
    numeric_cols = names(data)[sapply(data, is.numeric)],
    character_cols = names(data)[sapply(data, is.character)]
  )
  
  if (print_summary) {
    cat("=== Data Quality Report ===\n")
    cat("Number of rows:", results$n_rows, "\n")
    cat("Number of columns:", results$n_cols, "\n")
    cat("Duplicate rows:", results$duplicate_rows, "\n")
    cat("\nMissing values by column:\n")
    print(results$missing_values[results$missing_values > 0])
  }
  
  return(results)
}
```

### 5.2 Load Project Packages

```{r load-project-packages}
#' Load all R packages required for the project
#' 
#' @description Checks and installs (if needed) all R packages required for the project
#' @param install_missing Whether to install missing packages, default TRUE
#' @return No return value
#' @examples
#' load_project_packages()
load_project_packages <- function(install_missing = TRUE) {
  required_packages <- c(
    "tidyverse", "janitor", "skimr", "lubridate",
    "sf", "tmap", "plotly", "knitr", "car",
    "DBI", "RMySQL", "tidycensus", "tigris",
    "maptiles", "tidyterra", "leaflet", "gridExtra",
    "forcats", "viridis", "hrbrthemes", "scales"
  )
  
  for (pkg in required_packages) {
    if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
      if (install_missing) {
        cat("Installing", pkg, "...\n")
        install.packages(pkg)
        library(pkg, character.only = TRUE)
      } else {
        warning(paste("Package", pkg, "is not installed."))
      }
    }
  }
  
  cat("All required packages are loaded.\n")
}
```

## 6. Usage Examples

### 6.1 Complete Data Cleaning Workflow

```{r example-cleaning, eval=FALSE}
# Load packages
load_project_packages()

# Clean Airbnb data
df_cleaned <- clean_airbnb_data(file_path = "data/Airbnb_Open_Data_cleaned_first.csv")

# Check data quality
quality_report <- check_data_quality(df_cleaned)
```

### 6.2 Complete Visualization Workflow

```{r example-visualization, eval=FALSE}
# Plot borough distribution
plot_borough_distribution(df_cleaned)

# Plot room type distribution
plot_room_type_distribution(df_cleaned, by_borough = TRUE)

# Plot price distribution map (requires coordinate info)
plot_airbnb_prices(model_data)
```

### 6.3 Complete Modeling Workflow

```{r example-modeling, eval=FALSE}
# Prepare modeling data
model_data <- prepare_model_data("csv", csv_path = "data/model_data.csv")

# Build model
model <- lm(
  log_price ~ room_type + neighbourhood_group + minimum_nights +
              number_of_reviews + availability_365 + 
              median_income + median_rent + cancellation_policy,
  data = model_data
)

# Run diagnostics
run_model_diagnostics(model)
```

## 7. Function Index

| Function Name | Category | Description |
|:---|:---|:---|
| `clean_airbnb_data()` | Data Cleaning | Clean raw Airbnb data |
| `handle_missing_values()` | Data Cleaning | Fill missing values with median |
| `clean_date_column()` | Data Cleaning | Clean and standardize date column |
| `clean_minimum_nights()` | Data Cleaning | Clean minimum_nights column |
| `plot_income_map()` | Visualization | Plot income distribution map |
| `plot_airbnb_prices()` | Visualization | Plot price distribution map |
| `plot_borough_distribution()` | Visualization | Plot borough distribution |
| `plot_room_type_distribution()` | Visualization | Plot room type distribution |
| `run_model_diagnostics()` | Modeling | Run model diagnostics |
| `prepare_model_data()` | Modeling | Prepare modeling data |
| `spatial_join_airbnb_acs()` | Spatial Analysis | Spatial join Airbnb and ACS data |
| `create_wide_table()` | Spatial Analysis | Create wide format data |
| `check_data_quality()` | Utility | Check data quality |
| `load_project_packages()` | Utility | Load packages required for the project |

---

## Notes

1. Dependencies: All functions depend on specific R packages. Ensure required packages are installed before use. You can use load_project_packages() function to automatically install and load them.

2. Data Format: Ensure input data meets the format required by the functions. For example, spatial analysis functions require data to contain correct coordinate columns (lat, long) or geometry information.

3. Error Handling: Some functions contain basic error handling, but if issues arise, please check input data format and content.

4. Performance: For large datasets, some functions (especially spatial join functions) may take a longer time to run.
---





