---
title: "vis_finalproject"
author: "DYH"
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
# --- R Setup  ---

packages_to_install <- c(
  "tidyverse", # For data wrangling (dplyr, readr) and visualization (ggplot2)
  "sf",        # For geospatial data processing (R equivalent of geopandas)
  "tmap",      # For creating thematic maps (alternative to geoplot)
  "plotly",    # For interactive visualizations (alternative to plotly.express)
  "lubridate", # For handling dates
  "gridExtra", # For arranging multiple ggplot figures
  "knitr"      # For cleaner table rendering in reports
)

# Check, install, and load required packages
for (pkg in packages_to_install) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}

# Set default ggplot2 theme
theme_set(theme_minimal())

```
Introduction

This project uses R and R Markdown to visualize and engineer features from the 2020 NYC Airbnb dataset.
We explore relationships across boroughs, room types, and prices, and create geospatial visualizations to better understand spatial patterns in the data.
------------------------------------------------------------------------

### 1. Laod and Check data

Firstly, we read in data and clean data

```{r load-data, message=FALSE, warning=FALSE}
# read in data from CSV
data <- readr::read_csv('Data/airbnb_nyc_final_clean.csv')

# Display first 5 row
# Python: data.head()
knitr::kable(head(data))
```

```{r data-info}
# Check data types
# Python: data.info()
str(data)

# Python: data.describe()
summary(data)
```

```{r check-nulls}
# Check total missing values for each column
# Python: data.isnull().sum()
colSums(is.na(data))
```

```{r clean-data}
# Clean data:
# 1. Drop 'last_review' and 'host_name'
# 2. Replace NA in 'reviews_per_month' with 0

# Python:
# data.drop(['last_review','host_name'], axis=1, inplace=True)
# data['reviews_per_month'].fillna(0, inplace=True)

data_cleaned <- data %>%
  select(-last_review, -host_name) %>%
  mutate(reviews_per_month = tidyr::replace_na(reviews_per_month, 0))

data_cleaned <- data_cleaned %>%
  filter(neighbourhood_group != "Unresolved_API") %>%
  mutate(neighbourhood_group = factor(neighbourhood_group))

# Verify cleaning results
cat("Total missing values after cleaning:\n")
colSums(is.na(data_cleaned))
```

------------------------------------------------------------------------

### 2. Basic Categorical Analysis

We will visualize the distribution of listings across different categories,
such as boroughs and room types.

#### Listings by Borough

```{r plot-boroughs-styled, message=FALSE, warning=FALSE}
# Load 'forcats' for factor reordering (part of tidyverse)
library(forcats) 

# --- Fix (Robust Version) ---
# We now pre-calculate the counts and reorder the factor levels.
# This avoids the error by not using stat_count() inside the plot.

# 1. Calculate counts and reorder the 'neighbourhood_group' factor
#    'fct_reorder' sorts the factor 'neighbourhood_group' by the count 'n',
#    in descending order (.desc = TRUE).
borough_data_to_plot <- data_cleaned %>%
  count(neighbourhood_group, name = "n") %>%
  mutate(
    neighbourhood_group = fct_reorder(neighbourhood_group, n, .desc = TRUE)
  )

# 3. Plot using the new pre-counted data frame
#    --- IMPORTANT ---
#    We MUST use geom_col() now (not geom_bar).
#    We provide 'y = n' explicitly.
ggplot(borough_data_to_plot, aes(x = neighbourhood_group, y = n)) +
  
  geom_col(fill = "#56B4E9") + # Use geom_col() for pre-counted data
  
  # 1. All labels translated to English
  labs(title = "NYC Airbnb Listing Count by Borough",
       subtitle = "Manhattan and Brooklyn are the dominant boroughs",
       x = "Borough",
       y = "Number of Listings", # 'y' is now mapped to our count 'n'
       caption = "Source: NYC Airbnb Data (2020)") +
  
  # 2. Apply the built-in 'theme_minimal()'
  theme_minimal(base_family = "Arial") +
  
  # 3. Add extra tweaks to look professional (NYT style)
  theme(
    panel.grid.major.x = element_blank(), # Remove vertical grid lines
    panel.grid.minor = element_blank(),   # Remove all minor grid lines
    plot.title = element_text(face = "bold"), # Bold title
    axis.text.x = element_text(angle = 45, hjust = 1) 
  )
  
```

*The plot shows that Manhattan and Brooklyn have the highest number of listings among all boroughs*

#### Distribution of Listings by Room Type

```{r plot-room-types}
# Python (seaborn): sns.countplot(x='room_type', data=data_cleaned)
ggplot(data_cleaned, aes(x = room_type)) +
  geom_bar(fill = "darkorange") +
  labs(title = "Distribution of Listings by Room Type", x = "room_type", y = "listings")
```

*The plot indicates that most listings fall into two main categories: “Entire home/apt” and “Private room*

#### Detailed Room Type Composition Across Boroughs

```{r plot-stacked-borough-room-styled, message=FALSE, warning=FALSE}
# --- Styled Plot (English Labels + 'Minimal' Style) ---
############################################################################
# 1. (Important) We reuse the borough order computed in the previous code block
#    (borough_data_to_plot) to ensure that the x-axis order of this plot matches
#    the previous one (sorted by total listings)
#    We extract these levels and apply them to our plotting dataset

# Ensure that 'borough_data_to_plot' exists in the environment
if (!exists("borough_data_to_plot")) {
  # If it does not exist, recreate it (in case this block is run independently)
  borough_data_to_plot <- data_cleaned %>%
    count(neighbourhood_group, name = "n") %>%
    mutate(
      neighbourhood_group = fct_reorder(neighbourhood_group, n, .desc = TRUE)
    )
}

# Extract the ordered borough names
borough_order <- levels(borough_data_to_plot$neighbourhood_group)

# 2. Apply this ordering to the main dataset (for plotting purposes only)
data_cleaned_ordered <- data_cleaned %>%
  mutate(neighbourhood_group = factor(neighbourhood_group, levels = borough_order))

# 3. Plot
ggplot(data_cleaned_ordered, aes(x = neighbourhood_group, fill = room_type)) +
  
  geom_bar(position = "stack") +
  
  # 4. Convert all plot labels to English
  labs(title = "Room Type Composition by Borough",
       subtitle = "Manhattan is mostly 'Entire homes', others are more mixed",
       x = "Borough",
       y = "Number of Listings",
       fill = "Room Type", # this is legend
       caption = "Source: NYC Airbnb Data (2020)") +
  
  # 5. Apply theme_minimal() with custom adjustments
  theme_minimal(base_family = "Arial") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  
  # 6. Apply a more professional color palette 
  scale_fill_viridis_d(option = "cividis") # "mako", "viridis"
```

*This plot shows the composition of room types within each borough.*

------------------------------------------------------------------------

### 3. Geographic and Price Analysis

This is the core part of our analysis. We plot the listings on a map and examine their price distribution.

#### Top 10 Popular Neighborhoods

```{r plot-top10-neighbourhoods-styled, message=FALSE, warning=FALSE}
# --- Styled Plot (English Labels + 'Minimal' Style) ---

# 1. Pre-calculate the Top 10 data
#    (The user’s original code already implements this correctly)
top_10_neighbourhoods <- data_cleaned %>%
  count(neighbourhood, sort = TRUE) %>%
  top_n(10, n)

# 2. Plot
# We use fct_reorder to correctly order the x-axis
ggplot(top_10_neighbourhoods, aes(x = fct_reorder(neighbourhood, n, .desc = TRUE), y = n)) +
  
  # # Use a unified geom_col() style and apply a professional color palette
  geom_col(fill = "#56B4E9") + 
  
  # 3. # Translate all plot labels into English
  labs(title = "Top 10 Neighbourhoods by Listing Count",
       subtitle = "Specific areas like Williamsburg and Bedford-Stuyvesant dominate",
       x = "Neighbourhood",
       y = "Number of Listings",
       caption = "Source: NYC Airbnb Data (2020)") +
  
  # 4. # Apply theme_minimal() with custom styling adjustments
  theme_minimal(base_family = "Arial") +
  theme(
    panel.grid.major.x = element_blank(), 
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    # X轴标签需要倾斜，因为名称很长
    axis.text.x = element_text(angle = 45, hjust = 1) 
  )
```

#### Geospatial Distribution of Listings

```{r plot-geo-scatter-styled, message=FALSE, warning=FALSE}
# --- install and library packs ---
if (!require("maptiles")) {
  install.packages("maptiles")
  library(maptiles)
}
if (!require("sf")) {
  install.packages("sf")
  library(sf)
}
if (!require("tidyterra")) {
  install.packages("tidyterra")
  library(tidyterra)
}

# --- 1. Convert the dataset into an 'sf' spatial object ---
data_sf <- data_cleaned %>%
  filter(!is.na(lat), !is.na(long)) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)


# --- 2. automatically fetch map background tiles ---
nyc_map_tiles <- maptiles::get_tiles(
  data_sf, 
  provider = "CartoDB.PositronNoLabels"
)

# --- 3. Plotting: map as the background and points on top ---

# --- Fix ---
# 'maptiles::layer_maptiles()' is wrong
# This is the correct one !!! 'tidyterra::geom_spatraster_rgb()'

ggplot() +
  # First, draw the map background (using the correct function)
  tidyterra::geom_spatraster_rgb(data = nyc_map_tiles) +
  
  # Next, use geom_sf to plot our spatial data points
  geom_sf(data = data_sf, 
          aes(color = price), 
          alpha = 0.1, 
          size = 0.5) + 
  
  # Apply our previously defined color scale
  scale_color_viridis_c(trans = "log10", 
                        option = "magma",
                        alpha = 0.8) +
  
  # Translate to English version
  labs(title = "NYC Airbnb Listing Distribution and Price",
       subtitle = "Listings plotted over map, prices concentrated in Manhattan",
       x = "Longitude", 
       y = "Latitude",
       color = "Price (Log)",
       caption = "Source: NYC Airbnb Data (2020)") +
  
  # Apply 'NYT-style' theme
  theme_minimal(base_family = "Arial") +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  )
```

*This plot clearly outlines the shape of New York City. Higher-priced listings (yellow) are mainly concentrated in Manhattan*

# Interactive version available below
```{r plot-geo-scatter-interactive, message=FALSE, warning=FALSE}
# --- install and library 'leaflet' ---
# # Leaflet is the gold standard for building interactive maps in R
if (!require("leaflet")) {
  install.packages("leaflet")
  library(leaflet)
}
library(dplyr) # for %>%

# --- 1. Prepare data and color palette ---
# To create a meaningful legend, we define a price range.
# We use log10(price) to transform prices, similar to our ggplot version.
# Filter out zero or extremely high prices to make the visualization more meaningful.
data_for_map <- data_cleaned %>%
  filter(price > 0 & price <= 1000) %>%
  mutate(price_log = log10(price)) # ggplot 用的是 log10(price)

# 2. Create a color palette function (similar to scale_color_viridis)
# We apply the 'magma' palette to the log(price) values
pal <- colorNumeric(
  palette = "magma",
  domain = data_for_map$price_log
)

# --- 3. Create an interactive Leaflet map ---
# This will replace the static ggplot map
# You can zoom and pan freely in the RStudio viewer

leaflet(data_for_map) %>%
  
  # Add a clean “NYT-style” basemap (CartoDB Positron)
  addProviderTiles("CartoDB.PositronNoLabels") %>%
  
  # Add points (using CircleMarkers for better performance)
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = 1,          # Small circle markers
    stroke = FALSE,      # No border
    fillOpacity = 0.2,   # Set fill opacity
    color = ~pal(price_log), # Apply color palette
    
    # Create popup windows for hover/click events
    popup = ~paste(
      "<b>Name:</b>", name, "<br>",
      "<b>Price:</b> $", price, "<br>",
      "<b>Neighbourhood:</b>", neighbourhood, "<br>",
      "<b>Type:</b>", room_type
    )
  ) %>%

  # Add legend
  addLegend(
    pal = pal,
    values = ~price_log,
    opacity = 1,
    title = "Price (Log10)",
    position = "bottomright",
    # Customize legend tick labels to display actual prices (e.g., 100 instead of 2)
    labFormat = labelFormat(transform = function(x) round(10^x)) 
  )
```

```{r inspect-names}
# Check the actual column names in the data_cleaned dataframe
names(data_cleaned)

#### ighbourhood-level price choropleth map (Choropleth)
```

**Note: This choropleth requires the neighbourhoods.geojson file.**

```{r plot-choropleth, message=FALSE, warning=FALSE, results='hide'}

# 1. Load geographic boundary file
tryCatch({
  neighbourhoods_geo <- sf::st_read("neighbourhoods.geojson")
  
  # 2. Calculate median price by neighborhood
  price_by_neighbourhood <- data_cleaned %>%
    group_by(neighbourhood) %>%
    summarize(median_price = median(price, na.rm = TRUE))
  
  # # 3. Join the price data with the geographic boundaries
  map_data <- neighbourhoods_geo %>%
  left_join(price_by_neighbourhood, by = "neighbourhood")
  
  # 4. Use tmap to create a choropleth map (as an alternative to geoplot)
  tm <- tm_shape(map_data) +
    tm_polygons("median_price", 
                style = "quantile", 
                palette = "viridis",
                title = "median price ($)",
                popup.vars = c("neighbourhood", "median_price")) +
    tm_layout(title = "Median Airbnb Price Across NYC Neighborhoods")
  
  # Show Map
  tm
}, error = function(e) {
  cat("Error: Unable to load or process 'neighbourhoods.geojson'.\nPlease make sure the file is available in your working directory.\n", e$message)})
```

*If the `neighbourhoods.geojson` file is available, an interactive map will be displayed here*

#### Price Distribution





*The raw price data (left plot) is highly right-skewed. After applying a log transformation (right plot), the distribution becomes more normally shaped, which helps with both analysis and modeling.*

#### 	•	Interactive Geospatial Scatter Plot of Airbnb Listings (Plotly)
**dashboard interaction**


```{r}
 ###############################################################################################
library(plotly)
library(dplyr)

# To improve performance, we only sample 20,000 listings for the plot
data_sample <- data_cleaned %>% sample_n(min(20000, nrow(.)))

plot_ly(data_sample, 
        x = ~long, 
        y = ~lat, 
        color = ~room_type,  # colored by room type
        type = 'scattermapbox', 
        mode = 'markers',
        
        # --- English Labels (Tooltip Text) ---
        text = ~paste("Name:", name, "<br>Price: $", price, "<br>Type:", room_type),
        
        marker = list(size = 5, opacity = 0.5)) %>%
  
  # --- Plot Titles in English  ---
  plotly::layout(title = 'Interactive Map of NYC Listings by Room Type',
                 mapbox = list(style = 'carto-positron',
                               zoom = 9.5,
                               center = list(lon = -73.97, lat = 40.72)))
```







*This plot explores whether a listing’s popularity is related to its average price.*

#### Distribution of Minimum Nights


```{r plot-min-nights-styled, message=FALSE, warning=FALSE}
################################################################################################################

# We are focusing on listings with 30 nights or less
data_to_plot <- data_cleaned %>%
  filter(minimum_nights <= 30)

# --- Styled Plot (English Labels + 'Minimal' Style) ---

ggplot(data_to_plot, aes(x = minimum_nights)) +
  
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 30, 
                 fill = "#56B4E9", # A clean, light blue
                 alpha = 0.8) +
  
  geom_density(color = "#003366", # A dark blue line
               linewidth = 1) +
  
  labs(title = "Distribution of Minimum Nights",
       subtitle = "Focusing on listings requiring 30 nights or less",
       x = "Minimum Nights",
       y = "Density (proportion of listings)",
       caption = "Source: NYC Airbnb Data (2020)") +
  
  # 4. Use the built-in 'theme_minimal()'
  theme_minimal(base_family = "Arial") +
  
  # 5. Add extra tweaks to look professional
  theme(
    panel.grid.major.x = element_blank(), # Remove vertical grid lines
    panel.grid.minor = element_blank(),   # Remove minor grid lines
    plot.title = element_text(face = "bold") # Bold title
  )
```





